---
title: "Team pitching summaries"
author: "Martin Monkman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run Scoring Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Team pitching summaries: applying the principles of tidy evaluation


This vignette demonstrates how to call a function from within a `dplyr` pipe (using quasiquotation and quosures using `quo`) to count the number of pitchers used by each team in a single season.

First, we load the necessary packages: `Lahman` (containing the baseball data) and the data carpentry package `dplyr` (version 0.7.0 or higher):


```{r Setup, message=FALSE}

# package load 
library(Lahman)
library(dplyr)

```

#### Read and summarize the data

For this example, we'll use the data table `Teams` in the Lahman database.

Once it's loaded, the data are filtered and summarized using dplyr.

 - filters to select 2016,

Once filtered, a new variable indicating the number of relief appearances is calculated by subtracting the number of games started by the pitcher (GS) from the total number of games they appeared in (G).



```{r, message=FALSE}


data("Pitching")

Pitching_2016 <- Pitching %>%
  filter(yearID == "2016")

Pitching_2016 <- Pitching_2016 %>%
  mutate(GR = (G - GS))


View(Pitching_2016)


```

#### Summary table



```{r, message=FALSE}


```




***


#### Further reading:

* [dplyr 0.7.0](https://blog.rstudio.com/2017/06/13/dplyr-0-7-0/) (2017-06-13)

* [Programming with dplyr](http://dplyr.tidyverse.org/articles/programming.html)


**tidy evaluation**

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Tidy eval in ~5 minutes: <a href="https://t.co/mKBA7288jn">https://t.co/mKBA7288jn</a> — my latest attempt to cover the big ideas of tidy evaluation before you dive into the details <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a></p>&mdash; Hadley Wickham (@hadleywickham) <a href="https://twitter.com/hadleywickham/status/963147362410508288?ref_src=twsrc%5Etfw">February 12, 2018</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>


* [tidyeval resource roundup](https://maraaverick.rbind.io/2017/08/tidyeval-resource-roundup/) (2017-08-26)

* [Tidy evaluation](https://cran.r-project.org/web/packages/rlang/vignettes/tidy-evaluation.html)

* [Non-standard evaluation, how tidy eval builds on base R](https://edwinth.github.io/blog/nse/)


Bruno Rodrigues:

* [Functional programming and unit testing for data munging with R](https://b-rodrigues.github.io/fput/) (2017-08-25)

* [Why I find tidyeval useful](http://www.brodrigues.co/blog/2017-08-27-why_tidyeval/) (2017-08-27)

* [Lesser known dplyr 0.7* tricks](http://www.brodrigues.co/blog/2017-06-19-dplyr-0-70-tutorial/)


