---
title: "Team summaries"
author: "Martin Monkman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run Scoring Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Team summaries: Tidy evaluation


This vignette demonstrates how to call a function from within a `dplyr` pipe (using quasiquotation and quosures (`quo`)) to count the number of pitchers used by each team in a single season.

First, we load the necessary packages: `Lahman` (containing the baseball data) and the data carpentry package `dplyr` (version 0.7.0 or higher):


```{r, message=FALSE}

# package load 
library(Lahman)
library(dplyr)

```

#### Read and summarize the data

For this example, we'll use the data table `Teams` in the Lahman database.

Once it's loaded, the data are filtered and summarized using dplyr.

 - filters to select 2016,

Once filtered, a new variable indicating the number of relief appearances is calculated.



```{r, message=FALSE}


data("Pitching")

Pitching_2016 <- Pitching %>%
  filter(yearID == "2016")

Pitching_2016 <- Pitching_2016 %>%
  mutate(GR = (G - GS))


View(Pitching_2016)


```

#### Summary table



```{r, message=FALSE}


```




***


#### Further reading:

* [Tidy evaluation](https://cran.r-project.org/web/packages/rlang/vignettes/tidy-evaluation.html)

* [dplyr 0.7.0](https://blog.rstudio.com/2017/06/13/dplyr-0-7-0/) (2017-06-13)

* [Programming with dplyr](http://dplyr.tidyverse.org/articles/programming.html)

* [Why I find tidyeval useful](http://www.brodrigues.co/blog/2017-08-27-why_tidyeval/) (2017-08-27)

* [Lesser known dplyr 0.7* tricks](http://www.brodrigues.co/blog/2017-06-19-dplyr-0-70-tutorial/)

* [Non-standard evaluation, how tidy eval builds on base R](https://edwinth.github.io/blog/nse/)

